Reanalysis Step 1: Data Preparation

### 2025-08-14 reanalysis-1-data_preparation.qmd
dcsi

**Following the original analysis approach:**

1. **Glaucoma/Other Retina Conditions**: Missing values are recoded as 0 (absence of condition)
   - Clinical assumption: These conditions would be documented if present
   - This recoding happens BEFORE propensity score modeling

2. **CCI/DCSI**: Rows with missing values will be excluded (complete case)
   - These are key confounders that cannot be reliably imputed

@@

# Assess missing data AFTER recoding
missing_summary <- df_analysis |>
  dplyr::select(dplyr::all_of(c(ps_model_vars, "outcome_va_vi_binary"))) |>
  dplyr::summarise(dplyr::across(everything(), ~ sum(is.na(.)))) |>
  tidyr::pivot_longer(everything(), names_to = "variable", values_to = "n_missing") |>
  dplyr::mutate(
    pct_missing = round(n_missing / nrow(df_analysis) * 100, 2),
    var_type = dplyr::case_when(
      variable %in% c("outcome_va_vi_binary") ~ "Outcome",
      variable == "ever_lapse_binary" ~ "Exposure",
      variable %in% c("glaucoma_bef_hitplus_cat", "otherretina_bef_hitplus_cat") ~ "Covariate (recoded)",
      TRUE ~ "Covariate"
    ),
    handling = dplyr::case_when(
      variable %in% c("CCI", "DCSI") ~ "Exclude if missing",
      variable %in% c("glaucoma_bef_hitplus_cat", "otherretina_bef_hitplus_cat") ~ "Missing → 0 (no condition)",
      TRUE ~ "To be imputed"
    )
  ) |>
  dplyr::arrange(desc(n_missing))

@@

## Complete Case Analysis Dataset

```{r complete-cases}
# For comparison, create complete case dataset (no missing values)
df_complete <- df_analysis |>
  filter(!is.na(CCI) & !is.na(DCSI) & !is.na(baseline_VA_logMAR))

# Summary of exclusions
n_total <- nrow(df_analysis)
n_complete <- nrow(df_complete)
n_excluded <- n_total - n_complete

tibble(
  Dataset = c("Full dataset", "Complete cases", "Excluded"),
  N = c(n_total, n_complete, n_excluded),
  Percent = c(100, round(n_complete/n_total*100, 1), round(n_excluded/n_total*100, 1))
) |>
  kable(caption = "Complete case analysis summary")
```

@@

# Save the prepared dataset to network drive
saveRDS(df_analysis, file.path(reanalysis_data_dir, "df_analysis_prepared.rds"))
saveRDS(df_complete, file.path(reanalysis_data_dir, "df_complete_cases.rds"))

@@

```{r summary}
# Final summary
tibble(
  Metric = c(
    "Total observations",
    "Complete cases for main analysis", 
    "Missing CCI/DCSI",
    "Missing baseline VA",
    "Patients who lapsed",
    "Patients with vision impairment"
  ),
  N = c(
    nrow(df_analysis),
    nrow(df_complete),
    sum(is.na(df_analysis$CCI) | is.na(df_analysis$DCSI)),
    sum(is.na(df_analysis$baseline_VA_logMAR)),
    sum(df_analysis$ever_lapse_binary == 1, na.rm = TRUE),
    sum(df_analysis$outcome_va_vi_binary == 1, na.rm = TRUE)
  )
) |>
  kable(caption = "Dataset summary for reanalysis")
```

>>>>

**Modified approach for the reanalysis:**

1. **Glaucoma/Other Retina Conditions**: Missing values are recoded as 0 (absence of condition)
   - Clinical assumption: These conditions would be documented if present
   - This recoding happens BEFORE propensity score modeling
   - Maintains consistency with original analysis

2. **CCI/DCSI**: **NEW - Will be imputed** (change from original analysis)
   - Original analysis excluded rows with missing CCI/DCSI
   - Reanalysis will impute these to preserve sample size and reduce bias
   - These comorbidity indices can be reliably imputed from patient demographics and clinical characteristics

@@

# Assess missing data AFTER recoding
missing_summary <- df_analysis |>
  dplyr::select(dplyr::all_of(c(ps_model_vars, "outcome_va_vi_binary"))) |>
  dplyr::summarise(dplyr::across(everything(), ~ sum(is.na(.)))) |>
  tidyr::pivot_longer(everything(), names_to = "variable", values_to = "n_missing") |>
  dplyr::mutate(
    pct_missing = round(n_missing / nrow(df_analysis) * 100, 2),
    var_type = dplyr::case_when(
      variable %in% c("outcome_va_vi_binary") ~ "Outcome",
      variable == "ever_lapse_binary" ~ "Exposure",
      variable %in% c("glaucoma_bef_hitplus_cat", "otherretina_bef_hitplus_cat") ~ "Covariate (recoded)",
      TRUE ~ "Covariate"
    ),
    handling = dplyr::case_when(
      variable %in% c("glaucoma_bef_hitplus_cat", "otherretina_bef_hitplus_cat") ~ "Missing → 0 (no condition)",
      variable %in% c("CCI", "DCSI") ~ "To be imputed",
      TRUE ~ "To be imputed"
    )
  ) |>
  dplyr::arrange(desc(n_missing))

@@

## Dataset for Imputation

```{r imputation-dataset}
# Full dataset for imputation (no exclusions for missing CCI/DCSI)
df_for_imputation <- df_analysis

# For comparison, create complete case dataset (as in original analysis)
df_complete_original <- df_analysis |>
  filter(!is.na(CCI) & !is.na(DCSI))

# Summary of datasets
n_total <- nrow(df_analysis)
n_complete_original <- nrow(df_complete_original)
n_gained_by_imputation <- n_total - n_complete_original

tibble(
  Dataset = c(
    "Full dataset (for imputation)",
    "Complete cases (original approach)", 
    "Additional observations from imputation"
  ),
  N = c(n_total, n_complete_original, n_gained_by_imputation),
  Percent = c(
    100, 
    round(n_complete_original/n_total*100, 1), 
    round(n_gained_by_imputation/n_total*100, 1)
  )
) |>
  kable(caption = "Sample size comparison: Imputation vs complete case approach")
```

@@

# Save the prepared datasets to network drive
saveRDS(df_for_imputation, file.path(reanalysis_data_dir, "df_for_imputation.rds"))
saveRDS(df_complete_original, file.path(reanalysis_data_dir, "df_complete_original.rds"))

@@

```{r summary}
# Final summary
tibble(
  Metric = c(
    "Total observations",
    "Complete cases (original approach)", 
    "Missing CCI or DCSI (to be imputed)",
    "Missing baseline VA (to be imputed)",
    "Patients who lapsed",
    "Patients with vision impairment"
  ),
  N = c(
    nrow(df_for_imputation),
    nrow(df_complete_original),
    sum(is.na(df_for_imputation$CCI) | is.na(df_for_imputation$DCSI)),
    sum(is.na(df_for_imputation$baseline_VA_logMAR)),
    sum(df_for_imputation$ever_lapse_binary == 1, na.rm = TRUE),
    sum(df_for_imputation$outcome_va_vi_binary == 1, na.rm = TRUE)
  )
) |>
  kable(caption = "Dataset summary for reanalysis with imputation")
```
